---
title: "Lift — Congruence Motif (Mauerer 2022 STC) from Apache Helix"
output:
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Congruence Motif STC}
  %\VignetteEncoding{UTF-8}
---

```{r}
rm(list = ls())
seed <- 1
set.seed(seed)
```

```{r warning=FALSE, message=FALSE}
require(kaiaulu); require(data.table); require(stringi)
require(knitr); require(yaml); require(magrittr); require(igraph)
source("functions.R")
```

# Introduction

The `congruence_motif` model (Mauerer/Joblin/Tamburri/Paradis/Kazman/Apel
2022, IEEE TSE 48(8):3159-3184, doi:10.1109/TSE.2021.3082074) tests
the socio-technical congruence (STC) thesis using **graph motif
counting** rather than the Louvain community detection used by the
sibling `congruence` (smells-based) model. The two are independent
operationalizations of the same SE claim — *coordination gaps predict
defects* — and the framework's job is to triage which surface offers
the more falsifiable measurement.

Four motif templates (from kaiaulu `R/motif.R`):

```
Triangle:        dev─dev (comm edge) and both ─ same file (touch edge)        — positive
Anti-Triangle:   dev   dev (no comm) and both ─ same file                     — negative
Square:          dev─dev (comm) and dev─file1 and dev─file2 with file1─file2  — positive
Anti-Square:     dev   dev (no comm) across two dep-linked files              — negative
```

The motif-based congruence ratio:

```
congruence_motif = (Triangle + Square) / (Triangle + AntiTri + Square + AntiSq)
```

Method outline (parallels the smells-based congruence path, adds
dependency parsing):

1. Parse gitlog → `transform_gitlog_to_bipartite_network(mode="author-file")`.
2. Parse replies (mbox or GitHub) → `transform_reply_to_bipartite_network()`
   → `bipartite_graph_projection()` for dev↔dev edges.
3. Parse dependencies → `transform_dependencies_to_network(dependency_type="file")`.
4. `identity_match()` across both sources (REQUIRED — same cross-source
   bridge as the smells variant; without it, devs with mismatched
   mbox/git addresses double-count).
5. Merge all three networks into one combined graph; cast node colours
   to numeric so `igraph::count_subgraph_isomorphisms()` can use them.
6. For each motif in {triangle, anti_triangle, square, anti_square}:
   `motif_factory(name)` → `igraph::count_subgraph_isomorphisms(motif,
   network, method = "vf2")`. Emit four integer counts.

# Configuration

```{r}
tool <- parse_config("../tools.yml")
conf <- parse_config("../conf/helix.yml")

perceval_path <- get_tool_project("perceval", tool)
depends_path  <- get_tool_project("depends", tool)
git_repo_path <- get_git_repo_path(conf)
mbox_dir      <- "/Users/timm/Downloads/helix/mod_mbox/save_mbox_mail"
```

# Parse Git Log

```{r}
project_git <- parse_gitlog(perceval_path, git_repo_path) %>%
  filter_by_file_extension(get_file_extensions(conf), "file_pathname") %>%
  filter_by_filepath_substring(get_substring_filepath(conf),
                               "file_pathname")
project_git$author_datetimetz <- as.POSIXct(project_git$author_datetimetz,
                                            format = "%a %b %d %H:%M:%S %Y %z",
                                            tz = "UTC")
cat("git rows:", nrow(project_git),
    "  unique authors:", uniqueN(project_git$author_name_email), "\n")
```

# Parse Replies (Mbox)

```{r}
msgs <- parse_mbox_dir(perceval_path, mbox_dir)
cat("Mbox messages parsed:", nrow(msgs), "\n")
```

# Parse Dependencies

```{r}
# Depends produces <module>-file.json; kaiaulu parse_dependencies reads it.
# For Helix we point at helix-core's compiled class graph.
deps <- parse_dependencies("../data/helix/derived/depends/helix-core-file.json")
cat("dependency edges:", nrow(deps$edgelist), "\n")
```

# Identity Match (cross-source)

```{r}
project_log <- list(git_log = project_git, mbox = msgs)
project_log <- identity_match(
  project_log,
  name_column = c("author_name_email", "reply_from"),
  assign_identity_function = assign_exact_identity,
  use_name_only = FALSE,
  label = "identity_id"
)
project_git <- project_log$git_log
msgs        <- project_log$mbox
cat("Identities post-merge:",
    uniqueN(c(project_git$identity_id, msgs$identity_id)), "\n")
```

# Build the Three Networks

```{r}
# (1) author-file bipartite from gitlog
git_network <- transform_gitlog_to_bipartite_network(project_git,
                                                     mode = "author-file")

# (2) dev-dev from replies (project from dev-thread bipartite)
reply_network <- transform_reply_to_bipartite_network(msgs)
reply_network <- bipartite_graph_projection(
  reply_network, mode = TRUE,
  weight_scheme_function = weight_scheme_sum_edges)

# (3) file-file dep network
file_network <- transform_dependencies_to_network(deps,
                                                  dependency_type = "file")
```

# Combine Networks

```{r}
combined <- list()
combined[["nodes"]] <- unique(rbind(
  git_network[["nodes"]],
  reply_network[["nodes"]],
  file_network[["nodes"]]))
combined[["edgelist"]] <- rbind(
  git_network[["edgelist"]],
  reply_network[["edgelist"]],
  file_network[["edgelist"]])

g <- igraph::graph_from_data_frame(d = combined$edgelist,
                                   directed = FALSE,
                                   vertices = combined$nodes)
V(g)$color <- ifelse(V(g)$color == "black", 1, 2)
cat("merged graph: ", igraph::vcount(g), " nodes / ",
    igraph::ecount(g), " edges\n", sep = "")
```

# Count Motifs (Triangle, Anti-Triangle, Square, Anti-Square)

```{r}
count_motif <- function(motif_name, g) {
  m <- motif_factory(motif_name)
  img <- igraph::graph_from_data_frame(
    d = m$edgelist, directed = FALSE, vertices = m$nodes)
  V(img)$color <- ifelse(V(img)$color == "black", 1, 2)
  igraph::count_subgraph_isomorphisms(
    img, g, method = "vf2",
    edge.color1 = NULL, edge.color2 = NULL)
}

counts <- list(
  Triangle    = count_motif("triangle",      g),
  AntiTri     = count_motif("anti_triangle", g),
  Square      = count_motif("square",        g),
  AntiSquare  = count_motif("anti_square",   g)
)
print(unlist(counts))

ratio <- (counts$Triangle + counts$Square) /
         max(1, counts$Triangle + counts$AntiTri +
                counts$Square + counts$AntiSquare)
cat("congruence_motif ratio:", round(ratio, 3), "\n")
```

# Verdict on Helix

Mauerer 2022 reports the STC ratio as a per-snapshot summary. Higher
ratio = more dev-pair co-touches are accompanied by communication =
fewer realized coordination gaps. Helix's specific value goes into
the SD model's `comm_rate` calibration (replacing the `0.5` default).

The downstream SD model `paper/sd.py::congruence_motif()` uses
`comm_rate` as ctrl — flipping it from 0.2 to 0.8 reduces final
`Bugs`. At default the gap is `-68.78` (CONFIRM). Plugging Helix's
lifted ratio in via `paper/calibrate.py` would calibrate the
single-shot verdict against the real network's coordination signal.

# Outputs

```{r}
dir.create("../outputs", showWarnings = FALSE)
out <- data.table(
  project           = "helix",
  Triangle          = counts$Triangle,
  AntiTriangle      = counts$AntiTri,
  Square            = counts$Square,
  AntiSquare        = counts$AntiSquare,
  congruence_ratio  = round(ratio, 4),
  n_devs            = sum(V(g)$color == 1),
  n_files           = sum(V(g)$color == 2),
  n_edges           = igraph::ecount(g),
  seed              = seed
)
fwrite(out, "../outputs/lift_congruence_motif_helix.csv")
kable(out)
```

# Sanity Checks (Carlos's two-part)

**(1) Bug-count dependency**: not used by this lift. Anti-motif
counts are themselves the coordination-gap signal; Bugs is a
downstream consequence stock in the SD model, not a lifted quantity.

**(2) Identity bridging**: REQUIRED. Same `identity_match()` call as
the smells-based `congruence` lift, with two name_columns. Without
the merge, a developer appearing as `asebastian@linkedin` in mbox
and `asebastian@apache.org` in git would be treated as two nodes —
inflating anti-triangle counts and depressing the congruence ratio.

# Family-Member Test (3 projects with mbox + deps)

Pipeline-ready for Helix, airflow, tomcat (the three projects with
both mbox replies and a runnable Depends pass). Per-project runs
deferred; this notebook produces the Helix row first.

# References

- Mauerer, W., Joblin, M., Tamburri, D. A., Paradis, C., Kazman, R.,
  & Apel, S. (2022). In Search of Socio-Technical Congruence: A
  Large-Scale Longitudinal Study. *IEEE TSE* 48(8):3159-3184.
  doi:10.1109/TSE.2021.3082074.
- Cataldo, M., Wagstrom, P. A., Herbsleb, J. D., & Carley, K. M.
  (2006). Identification of Coordination Requirements: Implications
  for the Design of Collaboration and Awareness Tools. *CSCW '06*.
  (Original STC matrix-form.)
- Cordella, L. P., Foggia, P., Sansone, C., & Vento, M. (2004). A
  (sub)graph isomorphism algorithm for matching large graphs.
  *IEEE TPAMI* 26(10):1367-1372. (VF2.)
- kaiaulu R/motif.R — `motif_factory()` (triangle / anti_triangle /
  square / anti_square templates) + `transform_*_to_bipartite_network()`.
- kaiaulu vignettes/motif_analysis.Rmd — reference implementation
  walked through above.
- paper/sd.py::congruence_motif() — the SD model this lift calibrates.
- paper/sd.py::congruence() — the sibling smells-based model.
