---
title: "Explore — Workforce Cohort Helpers (Sterman learn)"
output:
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Explore Workforce Cohorts}
  %\VignetteEncoding{UTF-8}
---

```{r}
rm(list = ls()); seed <- 1; set.seed(seed)
```

```{r warning=FALSE, message=FALSE}
require(kaiaulu); require(data.table); require(stringi); require(magrittr); require(knitr)
```

# Introduction

Walks through Sterman-style cohort helpers in `R/myths_workforce.R`:

- `compute_cohorts()` — bucket each developer into Jr / Tr / Sr by
  tenure (years since first commit).
- `estimate_transition_rates()` — Jr→Tr / Tr→Sr promotion rates from
  sliding 90-day windows.

Downstream: `paper/sd.py:learn.train_rate`, `learn.promote_rate`.

# Project config

```{r warning=FALSE}
tool <- parse_config("../tools.yml")
conf <- parse_config("../../conf/helix.yml")
project_git <- parse_gitlog(get_tool_project("perceval", tool),
                            get_git_repo_path(conf))
project_git$author_datetimetz <- as.POSIXct(
  project_git$author_datetimetz,
  format = "%a %b %d %H:%M:%S %Y %z", tz = "UTC")
project_log <- identity_match(list(git_log = project_git),
                              c("author_name_email"),
                              assign_identity_function = assign_exact_identity,
                              use_name_only = FALSE,
                              label = "identity_id")
project_git <- project_log$git_log
```

# Step 1 — bucket developers by tenure

```{r}
cohorts <- compute_cohorts(project_git, jr_max_days = 365, tr_max_days = 1094)
kable(head(cohorts, 10))
cat("Jr / Tr / Sr counts:\n")
print(table(cohorts$cohort))
```

# Step 2 — estimate transition rates

```{r}
rates <- estimate_transition_rates(project_git, jr_max_days = 365,
                                   tr_max_days = 1094, window_days = 90)
kable(head(rates[order(window_start)], 8))
```

# Step 3 — annualised rates (the scalar fed to sd.py:learn)

```{r}
train_rate   <- median(rates$jr_to_tr_rate,   na.rm = TRUE) * (365 / 90)
promote_rate <- median(rates$tr_to_sr_rate,   na.rm = TRUE) * (365 / 90)
cat("train_rate (Jr -> Tr / yr):  ", round(train_rate,   3), "\n")
cat("promote_rate (Tr -> Sr / yr):", round(promote_rate, 3), "\n")
```

# What we learned

For Helix:
- ~24 distinct identities across ~10 years of history.
- Jr cohort dominates (most contributors are short-tenure drive-bys).
- Annualised Jr→Tr ≈ 0.81 — Sterman's workforce flow holds: most Jrs
  who DO stick around graduate within their first year.
- Earlier methodology used 365-day windows + 365-day Jr cutoff, which
  forced every Jr to graduate per window and saturated train_rate at
  1.0. Fixed to 90-day windows then annualised.
