---
title: "Explore — Refactor + Release-Snapshot Helpers"
output:
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Explore Refactor Snapshots}
  %\VignetteEncoding{UTF-8}
---

```{r}
rm(list = ls()); seed <- 1; set.seed(seed)
```

```{r warning=FALSE, message=FALSE}
require(kaiaulu); require(data.table); require(stringi); require(magrittr); require(knitr)
```

# Introduction

Walks through refactor + release helpers in `R/myths_refactor.R`:

- `get_release_tags()` — list git tags in lexicographic order.
- `get_tag_dates()` — pair tags with their commit timestamp.
- `flatten_refactoring_json()` — parse RefactoringMiner JSON output.
- `compute_pay_rate()` — fraction of commits per window containing
  ≥ 1 detected refactoring.
- `compute_born_rate_proxy()` — multi-file commit share (born-rate
  proxy).

Downstream: `paper/sd.py:debt.pay_rate`, `archpat.born_pat / born_leg`.

# Project config

```{r}
tool <- parse_config("../tools.yml")
conf <- parse_config("../../conf/helix.yml")
refminer_json <- "../../data/helix/derived/refminer.json"
cat("RefMiner JSON exists:", file.exists(refminer_json), "\n")
```

# Step 1 — list release tags

```{r}
tags <- get_release_tags(get_git_repo_path(conf))
kable(head(data.table(tag = tags), 10))
cat("total tags:", length(tags), "\n")
```

# Step 2 — tag dates

```{r}
tag_dates <- get_tag_dates(get_git_repo_path(conf))
kable(head(tag_dates, 8))
```

# Step 3 — parse RefMiner JSON

```{r}
refactorings <- flatten_refactoring_json(refminer_json)
kable(head(refactorings, 5))
cat("detected refactorings:", nrow(refactorings), "\n")
```

# Step 4 — pay rate (the scalar fed to debt.pay_rate)

```{r warning=FALSE}
project_git <- parse_gitlog(get_tool_project("perceval", tool),
                            get_git_repo_path(conf))
project_git$author_datetimetz <- as.POSIXct(
  project_git$author_datetimetz,
  format = "%a %b %d %H:%M:%S %Y %z", tz = "UTC")
pay <- compute_pay_rate(project_git, refactorings, window_days = 90)
kable(head(pay[order(window_start)], 8))
cat("median pay_rate:", round(median(pay$pay_rate, na.rm = TRUE), 3), "\n")
```

# Step 5 — born-rate proxy (multi-file commit share)

```{r}
born <- compute_born_rate_proxy(project_git, window_days = 90, min_files = 5)
kable(head(born[order(window_start)], 8))
cat("median born_rate proxy:",
    round(median(born$born_rate_proxy, na.rm = TRUE), 3), "\n")
```

# What we learned

For Helix:
- `pay_rate_median` ≈ 0.59 — most 90-day windows contain ≥ 1 RefMiner-
  detected refactoring. Sits comfortably inside debt.pay_rate's
  declared `[0, 1]` range. Helix is one of 5 Java projects whose
  pay_rate converges to 0.36–0.59 (F2).
- born-rate proxy (large multi-file commits) is rare — most commits
  touch ≤ 5 files. This is a CRUDE proxy; ideally we'd lift
  born_rate from new-feature commits identified via JIRA.
