---
title: "Explore — Mailing-List Reply Motifs (congruence)"
output:
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Explore Mailing Motifs}
  %\VignetteEncoding{UTF-8}
---

```{r}
rm(list = ls()); seed <- 1; set.seed(seed)
```

```{r warning=FALSE, message=FALSE}
require(kaiaulu); require(data.table); require(stringi); require(magrittr); require(knitr)
```

# Introduction

Walks through mailing-list helpers in `R/myths_mailing.R`:

- `parse_mbox_dir()` — wraps Perceval's mbox parser over a directory.
- `build_reply_edges()` — pivots (reply_id → identity) into a
  child→parent edge list.
- `detect_radio_silence()` — finds gaps where a previously-active
  boundary spanner stops replying.

Downstream: `paper/sd.py:congruence.broker_loss`, `Brokers_n`.

Note Carlos's review (PR #409): `detect_radio_silence` may overlap
with `R/smells.R:radio_silence` — to be folded in B6.

# Project config

```{r}
tool    <- parse_config("../tools.yml")
mbox_dir <- "../../data/helix/mbox"
cat("mbox dir exists:", dir.exists(mbox_dir), "\n")
```

# Step 1 — parse all mbox files

```{r}
perceval <- get_tool_project("perceval", tool)
msgs <- parse_mbox_dir(perceval, mbox_dir)
kable(head(msgs[, .(reply_id, in_reply_to_id, reply_from,
                    reply_datetimetz)], 5))
cat("total messages:", nrow(msgs), "\n")
```

# Step 2 — identity-merge senders

```{r}
m_list <- list(reply_log = msgs)
m_list <- identity_match(m_list, c("reply_from"),
                         assign_identity_function = assign_exact_identity,
                         use_name_only = FALSE,
                         label = "identity_id")
msgs <- m_list$reply_log
kable(head(unique(msgs[, .(identity_id, reply_from)]), 5))
```

# Step 3 — build reply edges

```{r}
edges <- build_reply_edges(msgs)
kable(head(edges, 10))
cat("edges:", nrow(edges), "\n")
```

# Step 4 — detect radio-silence (broker loss)

```{r}
silences <- detect_radio_silence(edges)
kable(head(silences, 10))
cat("radio-silence incidents:", nrow(silences), "\n")
```

# Headline stats

```{r}
brokers <- length(unique(c(silences$broker_id)))
clusters <- length(unique(c(edges$cluster_id)))
cat("brokers detected:        ", brokers, "\n")
cat("clusters identified:     ", clusters, "\n")
cat("longest silence (days):  ",
    if (nrow(silences)) max(silences$silence_days) else NA, "\n")
```

# What we learned

For Helix:
- 3 brokers across 5 Louvain clusters.
- 4 radio-silence incidents (1 broker disappeared, 3 weakened ties).
- The Newman/Cataldo broker-loss thesis (`congruence`) finds non-
  trivial mappings here.

Carlos's open question (B6): whether this implementation diverges
from `R/smells.R:radio_silence`. Side-by-side comparison TBD in a
follow-up commit.
