---
title: "Explore — Late-Hire Velocity Helpers (Brooks)"
output:
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Explore Late Hire Velocity}
  %\VignetteEncoding{UTF-8}
---

```{r}
rm(list = ls())
seed <- 1
set.seed(seed)
```

```{r warning=FALSE, message=FALSE}
require(kaiaulu)
require(data.table)
require(stringi)
require(magrittr)
require(knitr)
```

# Introduction

This notebook walks through the two helpers added in `R/myths_late_hire.R`
on Apache Helix:

- `detect_late_hires()` — finds developers whose first commit is
  ≥ N days after the project's first commit.
- `compute_velocity_changes()` — for each late hire, computes the
  pre/post commit rate of *veterans* (devs who joined before).

The downstream `brooks_tax_median = (pre - post) / pre` is the
calibrated metric consumed by the SD model `paper/sd.py:brooks`.

# Project Configuration File

```{r}
tool <- parse_config("../tools.yml")
conf <- parse_config("../../conf/helix.yml")
```

`conf` carries the gitlog path + filter list. `tool` carries paths to
Perceval, RefactoringMiner, etc.

# Step 1 — parse the gitlog via Perceval

```{r warning=FALSE}
project_git <- parse_gitlog(
  get_tool_project("perceval", tool),
  get_git_repo_path(conf)
) %>%
  filter_by_file_extension(get_file_extensions(conf), "file_pathname") %>%
  filter_by_filepath_substring(get_substring_filepath(conf), "file_pathname")

# Perceval emits "Tue Jun 21 18:56:46 2011 -0700"; parse to POSIXct
project_git$author_datetimetz <- as.POSIXct(
  project_git$author_datetimetz,
  format = "%a %b %d %H:%M:%S %Y %z", tz = "UTC"
)

kable(head(project_git[, .(commit_hash, author_name_email,
                           author_datetimetz, file_pathname)], 5))
```

# Step 2 — merge developer aliases

```{r}
project_log <- list(git_log = project_git)
project_log <- identity_match(project_log, c("author_name_email"),
                              assign_identity_function = assign_exact_identity,
                              use_name_only = FALSE,
                              label = "identity_id")
project_git <- project_log$git_log

# 5 example identity rows
kable(head(unique(project_git[, .(identity_id, author_name_email)]), 5))
```

# Step 3 — detect late hires

```{r}
late_hires <- detect_late_hires(project_git, min_project_age_days = 365)
cat("late hires detected:", nrow(late_hires), "\n")
kable(head(late_hires[order(days_after_project_start)], 10))
```

# Step 4 — compute pre/post velocity per hire

```{r}
window_days <- 90
vc <- compute_velocity_changes(project_git, late_hires, window_days)
vc[, brooks_tax := (pre_velocity - post_velocity) / pre_velocity]
kable(head(vc[order(-brooks_tax)], 10))
```

# Headline stats

```{r}
cat("median brooks_tax:", median(vc$brooks_tax, na.rm = TRUE), "\n")
cat("mean   brooks_tax:", mean(vc$brooks_tax,   na.rm = TRUE), "\n")
cat("n_hires:           ", nrow(vc), "\n")
```

# What we learned

For Helix specifically: 65 late-hire events over the project lifetime;
median brooks_tax ≈ 0.11 (~11% velocity drop on veterans in the 90-day
window after each new hire). The scalar feeds the SD model's
verification machinery — see `paper/cross_project.py` for the
8-project comparison.
