---
title: "Lift — Brooks Model from Apache Helix"
output: 
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Lift Brooks}
  %\VignetteEncoding{UTF-8}
---

```{r}
rm(list = ls())
seed <- 1
set.seed(seed)
```

```{r warning=FALSE, message=FALSE}
require(kaiaulu)
require(data.table)
require(stringi)
require(knitr)
require(yaml)
require(magrittr)
source("functions.R")
```

# Introduction

The Brooks model expresses the thesis from *The Mythical Man-Month*:
adding people to a late project makes it later. The SD form has two
stocks (veteran developers and new hires) and a productivity tax on
veterans when newcomers join (training time, comm overhead).

This notebook lifts Brooks parameters from Apache Helix using only
the git log and kaiaulu's identity_match. No proprietary tools needed.

The lift produces three calibration values:
- `late_hire_events`: timestamps where new committers first appear
- `pre_velocity`: veteran commit rate in the window before each hire
- `post_velocity`: veteran commit rate in the post-hire window

The SD parameter `brooks_tax` = (pre_velocity - post_velocity) /
pre_velocity. Positive values support Brooks; negative values
falsify (paradoxical: new hires accelerate veterans).

# Project Configuration

```{r}
tool <- parse_config("../tools.yml")
conf <- parse_config("../conf/helix.yml")

perceval_path <- get_tool_project("perceval", tool)
git_repo_path <- get_git_repo_path(conf)

file_extensions      <- get_file_extensions(conf)
substring_filepath   <- get_substring_filepath(conf)
```

# Parse Git Log

```{r}
project_git <- parse_gitlog(perceval_path, git_repo_path)

project_git <- project_git %>%
  filter_by_file_extension(file_extensions, "file_pathname") %>%
  filter_by_filepath_substring(substring_filepath, "file_pathname")

project_git$author_datetimetz <- as.POSIXct(project_git$author_datetimetz,
                                            format = "%a %b %d %H:%M:%S %Y %z",
                                            tz = "UTC")
```

Sanity check the basic shape:

```{r}
nrow(project_git)
length(unique(project_git$commit_hash))
length(unique(project_git$author_name_email))
```

# Identity Resolution

Merge name+email aliases so one human contributor is one developer.
Required by Carlos's sanity check #2 (single source here, but
identity_match is the canonical kaiaulu step).

```{r}
project_log <- list(git_log = project_git)
name_column <- c("author_name_email")
project_log <- identity_match(project_log,
                              name_column,
                              assign_identity_function = assign_exact_identity,
                              use_name_only = FALSE,
                              label = "identity_id")
project_git <- project_log$git_log
```

After identity_match, each row has `identity_id` and `raw_name`. Use
`identity_id` for all downstream cohort analysis.

# Detect Late-Hire Events

A late hire is a developer whose first commit is at least
`min_project_age_days` after project start. Function defined in
`functions.R`:

```{r}
project_start <- min(project_git$author_datetimetz)
late_hires    <- detect_late_hires(project_git,
                                   min_project_age_days = 365)
kable(head(late_hires, 20))
```

# Compute Veteran Velocity Before and After Each Hire

Window size in days:

```{r}
window_days <- 90
```

```{r}
velocity_changes <- compute_velocity_changes(project_git,
                                             late_hires,
                                             window_days = window_days)
kable(velocity_changes)
```

# Compute brooks_tax

```{r}
brooks_tax_values <- velocity_changes[,
  brooks_tax := (pre_velocity - post_velocity) / pre_velocity
]

summary(brooks_tax_values$brooks_tax)
```

Median brooks_tax for Helix is the calibration value for the SD
`brooks` model. A positive median supports Brooks; a negative or near-zero
median is itself an interesting falsification.

# Sanity Checks

Carlos's two-part check:

**(1) Bug-count dependency**: this lift does NOT use bug count, so
the choice of communication source (mbox/github/jira) is irrelevant.
The lift would work on any git-only project.

**(2) Identity bridging**: we used identity_match on a single source
(git). The lift does not require comms-to-code bridging. If the
brooks model were later extended to discriminate bug-introducing
late-hires from feature-introducing late-hires, JIRA + git
identity_match would be required, and GitHub would need an extra
alias source.

# Outputs for the SD Framework

Write the calibrated `brooks_tax` value(s) to a CSV the Python
framework can consume:

```{r}
out <- data.table(
  project          = "helix",
  brooks_tax_median = median(brooks_tax_values$brooks_tax, na.rm = TRUE),
  brooks_tax_mean   = mean(brooks_tax_values$brooks_tax,   na.rm = TRUE),
  n_hires          = nrow(velocity_changes),
  window_days      = window_days,
  seed             = seed
)
fwrite(out, "../outputs/lift_brooks_helix.csv")
kable(out)
```
