---
title: "Lift — Learn Model from Apache Helix"
output:
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Lift Learn}
  %\VignetteEncoding{UTF-8}
---

```{r}
rm(list = ls())
seed <- 1
set.seed(seed)
```

```{r warning=FALSE, message=FALSE}
require(kaiaulu); require(data.table); require(stringi)
require(knitr); require(yaml); require(magrittr)
source("functions.R")
```

# Introduction

The `learn` model (Sterman 2000) tracks a workforce-flow pipeline:
Jr → Tr → Sr → Ment(or). `ctrl` is `Sr` (senior stock). Thesis:
removing seniors (`Sr = 0`) starves the training pipeline.

This notebook lifts cohort counts + transition rates from Apache Helix.
Only git + kaiaulu's identity_match required.

# Project Configuration

```{r}
tool <- parse_config("../tools.yml")
conf <- parse_config("../conf/helix.yml")
git_repo_path <- get_git_repo_path(conf)
perceval_path <- get_tool_project("perceval", tool)
```

# Parse Git Log + Identity Match

```{r}
project_git <- parse_gitlog(perceval_path, git_repo_path) %>%
  filter_by_file_extension(get_file_extensions(conf), "file_pathname") %>%
  filter_by_filepath_substring(get_substring_filepath(conf), "file_pathname")
project_git$author_datetimetz <- as.POSIXct(project_git$author_datetimetz,
                                            format = "%a %b %d %H:%M:%S %Y %z",
                                            tz = "UTC")
project_git <- identity_match(list(git_log = project_git),
                              c("author_name_email"),
                              assign_identity_function = assign_exact_identity,
                              use_name_only = FALSE,
                              label = "identity_id")$git_log
uniqueN(project_git$identity_id)
```

# Compute Cohorts at End of Project

```{r}
cohorts <- compute_cohorts(project_git,
                           jr_max_days = 365,
                           sr_min_days = 1095)
kable(head(cohorts[order(-tenure_days)], 20))
table(cohorts$cohort)
```

# Estimate Transition Rates

```{r}
rates <- estimate_transition_rates(project_git,
                                   jr_max_days = 365,
                                   sr_min_days = 1095)
str(rates)
```

# Outputs

```{r}
dir.create("../outputs", showWarnings = FALSE)
counts <- table(cohorts$cohort)
out <- data.table(
  project       = "helix",
  Jr_n          = as.integer(counts["Jr"]),
  Tr_n          = as.integer(counts["Tr"]),
  Sr_n          = as.integer(counts["Sr"]),
  train_rate    = rates$train_rate,
  promote_rate  = rates$promote_rate,
  n_slices      = rates$n_slices,
  seed          = seed
)
fwrite(out, "../outputs/lift_learn_helix.csv")
kable(out)
```

# Verdict Interpretation

`learn`'s thesis (Sr=0 starves training) is structural — it requires
running the SD model with `Sr` set to 0 and comparing to the calibrated
Sr count from this lift. Plug the lifted Jr_n/Tr_n/Sr_n into the model
init and rerun `model.rq(bg=calib)`.
