---
title: "Lift — Archpat Model from Apache Helix (GoF + RefactoringMiner)"
output: 
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Lift Archpat}
  %\VignetteEncoding{UTF-8}
---

```{r}
rm(list = ls())
seed <- 1
set.seed(seed)
```

```{r warning=FALSE, message=FALSE}
require(kaiaulu)
require(data.table)
require(stringi)
require(knitr)
require(yaml)
require(magrittr)
source("functions.R")
```

# Introduction

The archpat model (Ric Kazman's claim) tests whether design patterns
applied to an already-bad codebase can pay down architectural debt.
The model has three stocks (Patterned files, Legacy files, Drift
files) plus a Debt accumulator.

This notebook lifts archpat parameters from Apache Helix using:
- **pattern4.jar** for GoF pattern instances per file (Patterned partition)
- **RefactoringMiner** for refactoring events (gen_pat: pattern application rate)
- **kaiaulu gitlog + JIRA** for the Debt accumulator (defect frequency)
- **Arcan** (fallback) — see note at end if pattern4 is unavailable

Parameters lifted: 11 of 15 archpat init keys. Three (intr_rate,
drift_to_legacy, pay_rate) remain fitted via sd.opt() because they
appear only as net Debt change, not as observable individual flows.
One (pat_strength) is dead code — flagged for Ric.

# Project Configuration

```{r}
tool <- parse_config("../tools.yml")
conf <- parse_config("../conf/helix.yml")

perceval_path     <- get_tool_project("perceval", tool)
pattern4_jar      <- get_tool_project("pattern4", tool)
refminer_jar      <- get_tool_project("refactoring_miner", tool)

git_repo_path     <- get_git_repo_path(conf)
jira_issues_path  <- get_jira_issues_path(conf, "project_key_1")
issue_id_regex    <- get_issue_id_regex(conf)

file_extensions      <- get_file_extensions(conf)
substring_filepath   <- get_substring_filepath(conf)
```

# Parse Git Log

```{r}
project_git <- parse_gitlog(perceval_path, git_repo_path)

project_git <- project_git %>%
  filter_by_file_extension(file_extensions, "file_pathname") %>%
  filter_by_filepath_substring(substring_filepath, "file_pathname")

project_git$author_datetimetz <- as.POSIXct(project_git$author_datetimetz,
                                            format = "%Y-%m-%d %H:%M:%S")

# Mutate gitlog to add commit_message_id (extracts issue keys from commit messages)
project_git <- parse_commit_message_id(project_git, issue_id_regex)
```

# Parse JIRA

```{r}
parsed_jira    <- parse_jira(jira_issues_path)
jira_issues    <- parsed_jira$issues
jira_comments  <- parsed_jira$comments

# CORRECTION: parse_jira()'s issue_status is statusCategory.name
# ("To Do" / "In Progress" / "Done"), NOT status.name. Use "Done"
# for "completed" semantics. See kaiaulu_notes/known_bugs.md.
jira_bugs_done <- jira_issues[issue_type == "Bug" & issue_status == "Done"]

nrow(jira_bugs_done)
```

# GoF Pattern Detection (Patterned Partition)

Run pattern4.jar on a snapshot of Helix source (e.g. the HEAD commit
of each release tag). Function defined in functions.R; calls
parse_gof_patterns wrapper:

```{r}
release_tags <- get_release_tags(git_repo_path)
gof_per_tag  <- lapply(release_tags, function(tag) {
  source_path <- checkout_snapshot(git_repo_path, tag)
  parse_gof_patterns(pattern4_jar, source_path)
})
names(gof_per_tag) <- release_tags
```

`parse_gof_patterns` returns a data.table per snapshot with:
```
file_path        chr   path within the snapshot
pattern_type     chr   "Singleton" / "Factory Method" / ...
pattern_role     chr   role of this file in the pattern instance
```

Build the Patterned partition for the latest snapshot:

```{r}
latest_gof  <- gof_per_tag[[length(gof_per_tag)]]
patterned_files <- unique(latest_gof$file_path)
length(patterned_files)
```

# RefactoringMiner (gen_pat — Pattern Application Rate)

```{r}
refminer_output <- run_refactoring_miner(refminer_jar, git_repo_path)
refactorings    <- flatten_refactoring_json(refminer_output)
```

`flatten_refactoring_json` (in functions.R) converts the nested JSON
into a flat data.table:
```
commit_hash, refactoring_type, refactoring_description,
left_locations, right_locations
```

Per-month refactor rate as proxy for gen_pat:

```{r}
refactorings$commit_datetimetz <- merge_commit_dates(refactorings,
                                                     project_git)
refactor_rate_per_month <- refactorings[,
  .N,
  by = .(month = format(commit_datetimetz, "%Y-%m"))
]
gen_pat_helix <- mean(refactor_rate_per_month$N) / 30  # per day
gen_pat_helix
```

# Build Legacy and Drift Partitions

A file is **Legacy** if it has high accumulated bug frequency AND no
GoF pattern participation. A file is **Drift** if it has changed
hands recently (high churn) but has no pattern instance yet.

```{r}
file_bug_freq <- compute_file_bug_frequency(project_git,
                                            jira_bugs_done,
                                            issue_id_regex)
file_churn    <- compute_file_churn(project_git,
                                    window_days = 180)

partition <- assign_file_partition(
  patterned_files = patterned_files,
  file_bug_freq   = file_bug_freq,
  file_churn      = file_churn,
  legacy_bug_threshold   = 5,
  drift_churn_threshold  = 0.7
)

partition[, .N, by = stock]
```

# Lift the Other archpat Parameters

```{r}
init_archpat <- list(
  Patterned    = partition[stock == "Patterned",   .N],
  Legacy       = partition[stock == "Legacy",      .N],
  Drift        = partition[stock == "Drift",       .N],
  Debt         = sum(file_bug_freq$bug_count),  # total accumulated debt
  Feat         = compute_feature_velocity(project_git, jira_issues),
  migrate      = compute_migration_rate(gof_per_tag),
  decay_rate   = compute_smell_appearance_rate(gof_per_tag),
  gen_pat      = gen_pat_helix,
  gen_leg      = compute_legacy_growth_rate(project_git, patterned_files),
  born_pat     = compute_born_pat_rate(refactorings, gof_per_tag),
  born_leg     = compute_born_leg_rate(project_git, patterned_files)
  # intr_rate, drift_to_legacy, pay_rate: fitted via sd.opt(),
  # not lifted (only net Debt change is observable)
  # pat_strength: defined in sd.py but unused in step() — flag for Ric
)
str(init_archpat)
```

# Sanity Checks

**(1) Bug-count dependency**: this lift uses bug counts heavily for
the Legacy partition and the Debt accumulator. Therefore JIRA is
required. Helix has JIRA — OK. Datasets without JIRA (or equivalent
issue tracker) cannot support archpat as-is.

**(2) Identity bridging**: comms data (mbox, github) is not used in
this lift, so the comms↔code identity_match constraint does not
apply. Only git's own identity_match runs.

# Outputs for the SD Framework

```{r}
out <- as.data.table(init_archpat)
out[, project := "helix"]
out[, seed := seed]
fwrite(out, "../outputs/lift_archpat_helix.csv")
kable(out)
```

# Fallback: Arcan Instead of pattern4.jar

If pattern4.jar is unavailable, substitute Arcan smells for the
Patterned/Legacy partition. The smells are conceptually different
(anti-patterns rather than GoF patterns) but structurally identical
— each row is (file, smell_type, smell_id). The downstream
partition assignment in `functions.R::assign_file_partition` accepts
either Arcan or pattern4 output as long as it has a `file_path`
column.

Note in any paper deliverable which detector was used; the model is
agnostic but the empirical numbers will differ.
